\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks

### 2015-09-17 nnt_manu.Rnw
added refs for neural network packages, need to mention others in the RSNNS paper

<<eval = F, echo = F>>=
# total downloads for neural network packages off CRAN
# make sure to change cdate
library(cranlogs)

tots <- c(
  sum(cran_downloads('nnet', from = '2000-01-01', to = '2015-09-13')$count),
  sum(cran_downloads('neuralnet', from = '2000-01-01', to = '2015-09-13')$count),
  sum(cran_downloads('RSNNS', from = '2000-01-01', to = '2015-09-13')$count),
  sum(cran_downloads('FCNN4R', from = '2000-01-01', to = '2015-09-13')$count)
)
@

Packages available in \proglang{R} to create \ac{mlp} neural networks include \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}. The \pkg{train} can be used to create a neural network model by repeated training with the \code{nnet}method available from \pkg{nnet}.  An additional package, \pkg{FCNN4R}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.  At the time of writing, \pkg{FCNN4R} is a new package with less than 1\% of the total downloads of all neural network packages on CRAN.  As such, \pkg{NeuralNetTools} does not include methods for the \pkg{FCNN4R} package, although future versions may do so depending on the popularity of \pkg{FCNN4R}.  In general, the remaining packages use similar methods to create \ac{mlp} networks such that the resulting models are comparable.  Differences between the packages primarily include default values of the arguments and the ability to include multiple response variables or multiple hidden layers.  The primary functions in each package also vary in the type of input used to specify the model structure.  Differences that influence how the \pkg{NeuralNetTools} functions handle object classes from the different packages are described below.

@@

A sample dataset is also included with \pkg{NeuralNetTools}.  The \code{neuraldat} dataset is a simple \code{data.frame} with 300 rows of observations and columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}.  The input variables are random observations from a standard normal distribution and the response variables are linear combinations of the input variables with additional random components.  The response variables are also standardized from zero to one.  A common approach for data preprocessing prior to creating a nueral network is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.  The \code{neuraldat} package provides data following this structure.  Novel datasets can be preprocessed in the common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.

@@

The number of existing functions in \pkg{R} to view neural networks is minimal.  Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance. To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks in R.  Although useful for viewing the basic structure, the output is visually minimal and does not include extensive options for customization.

Tbe \code{plotnet} function plots a neural network as a neural interpretation diagram \citep{Ozesmi99} and includes many options to customize plot aesthetics. A neural interpration diagram is a modification to the standard conceptual illustration of the \ac{mlp} network that allows the user to change the thickness and color of the weight connections based on mangnitude and sign, respecitively.  The default settings plot positive weights between layers as black lines and negative weights as grey lines. Line thickness is in proportion to relative magnitude of each weight. The first layer includes only input variables with nodes labelled arbitrarily as I1 through In for n input variables. One through many hidden layers are plotted with each node in each layer labelled as H1 through Hn. The output layer is plotted last with nodes labeled as O1 through On. Bias nodes connected to the hidden and output layers are also shown. Differences in the behavior of \code{plotnet} for different object classes are minimal, with the exception of \code{mlp} class networks that do not include bias layers.

>>>>

<<eval = F, echo = F>>=
# total downloads for neural network packages off CRAN
# make sure to change cdate
library(cranlogs)
today <- as.character(as.Date(Sys.time()))

tots <- c(
  sum(cran_downloads('nnet', from = '2000-01-01', to = today)$count),
  sum(cran_downloads('neuralnet', from = '2000-01-01', to = today)$count),
  sum(cran_downloads('RSNNS', from = '2000-01-01', to = today)$count),
  sum(cran_downloads('FCNN4R', from = '2000-01-01', to = today)$count),
  sum(cran_downloads('AMORE', from = '2000-01-01', to = today)$count),
  sum(cran_downloads('nonmlp', from = '2000-01-01', to = today)$count),
  sum(cran_downloads('qrnn', from = '2000-01-01', to = today)$count),
  sum(cran_downloads('frbf', from = '2000-01-01', to = today)$count)
)

@

See refs in the RSNNS paper!!

Methods in \pkg{NeuralNetTools} were written for several \proglang{R} packages that can be used to create \ac{mlp} neural networks: \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}. The \code{train} function from the \pkg{caret} package \citep{Kuhn15} can be used to create a neural network model by repeated training with functions from the remaining packages.  Additional packages without \pkg{NeuralNetTools} methods include....

An additional package, \pkg{FCNN4R} \citep{Klima15}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.  At the time of writing, \pkg{FCNN4R} is a new package with less than 1\% of the total downloads of all neural network packages on CRAN.  As such, \pkg{NeuralNetTools} does not include methods for the \pkg{FCNN4R} package, although future versions may do so depending on the popularity of \pkg{FCNN4R}.  In general, the remaining packages use similar methods to create \ac{mlp} networks such that the resulting models are comparable.  Differences between the packages primarily include default values of the arguments and the ability to include multiple response variables or multiple hidden layers.  The primary functions in each package also vary in the type of input used to specify the model structure.  Differences that influence how the \pkg{NeuralNetTools} functions handle object classes from the different packages are described below.

@@

A sample dataset is also included with \pkg{NeuralNetTools}.  The \code{neuraldat} dataset is a simple \code{data.frame} with 300 rows of observations and columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}.  The input variables are random observations from a standard normal distribution and the response variables are linear combinations of the input variables with additional random components.  The response variables are also standardized from zero to one.  A common approach for data preprocessing prior to creating a nueral network is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.  Novel datasets can be preprocessed to this common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.

@@

The number of existing functions in \pkg{R} to view neural networks is minimal.  Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance. To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks in R.  Although useful for viewing the basic structure, the output is minimal and does not include extensive options for customization.

Tbe \code{plotnet} function plots a neural network as a neural interpretation diagram \citep{Ozesmi99} and includes several options to customize plot aesthetics. A neural interpration diagram is a modification to the standard conceptual illustration of the \ac{mlp} network that allows the user to change the thickness and color of the weight connections based on mangnitude and sign, respectively.  The default settings plot positive weights between layers as black lines and negative weights as grey lines. Line thickness is in proportion to absolte magnitude of each weight. The first layer includes only input variables with nodes labelled arbitrarily as I1 through In for n input variables. One through many hidden layers are plotted with each node in each layer labelled as H1 through Hn. The output layer is plotted last with nodes labeled as O1 through On. Bias nodes connected to the hidden and output layers are also shown. Differences in the behavior of \code{plotnet} for different object classes are minimal, with the exception of \code{mlp} class networks that do not include bias layers.  The following shows some examples of basic plots

### 2015-09-17 nnt_manu.tex
added refs for neural network packages, need to mention others in the RSNNS paper

Packages available in \proglang{R} to create \ac{mlp} neural networks include \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}. The \pkg{train} can be used to create a neural network model by repeated training with the \code{nnet}method available from \pkg{nnet}.  An additional package, \pkg{FCNN4R}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.  At the time of writing, \pkg{FCNN4R} is a new package with less than 1\% of the total downloads of all neural network packages on CRAN.  As such, \pkg{NeuralNetTools} does not include methods for the \pkg{FCNN4R} package, although future versions may do so depending on the popularity of \pkg{FCNN4R}.  In general, the remaining packages use similar methods to create \ac{mlp} networks such that the resulting models are comparable.  Differences between the packages primarily include default values of the arguments and the ability to include multiple response variables or multiple hidden layers.  The primary functions in each package also vary in the type of input used to specify the model structure.  Differences that influence how the \pkg{NeuralNetTools} functions handle object classes from the different packages are described below.

>>>>

Packages available in \proglang{R} to create \ac{mlp} neural networks include \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}. The \code{train} function from the \pkg{caret} package \citep{Kuhn15} can be used to create a neural network model by repeated training with functions from the remaining packages.  An additional package, \pkg{FCNN4R} \citep{Klima15}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.  At the time of writing, \pkg{FCNN4R} is a new package with less than 1\% of the total downloads of all neural network packages on CRAN.  As such, \pkg{NeuralNetTools} does not include methods for the \pkg{FCNN4R} package, although future versions may do so depending on the popularity of \pkg{FCNN4R}.  In general, the remaining packages use similar methods to create \ac{mlp} networks such that the resulting models are comparable.  Differences between the packages primarily include default values of the arguments and the ability to include multiple response variables or multiple hidden layers.  The primary functions in each package also vary in the type of input used to specify the model structure.  Differences that influence how the \pkg{NeuralNetTools} functions handle object classes from the different packages are described below.
